---
title: "Pipeline Run & Data Check"
author: "Keithley Flores — STA 160 Final Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Purpose

This notebook is **not** the report. It is the one-time (well, run-it-whenever-
the-data-changes) setup step that:

1. sources `config.R` and the feature scripts `01`–`04`,
2. runs the full extraction loop over all selected species (`05_run_all.R`),
3. writes `output/features.csv`, and
4. sanity-checks that the CSV really contains all the species you expect.

Knit or run this first. **Only after** `output/features.csv` shows ~375 rows
across 5 species should you knit `final_report.Rmd` — that report just reads
the CSV this notebook produces.

> Why separate files? The report was rendering "1 species / 2 images" because
> a stale 2-row `features.csv` was sitting in `output/` from an earlier
> single-leaf test. Keeping the *run* (this file) apart from the *report*
> makes it obvious whether the data has actually been regenerated.

# Locate the project scripts

Set this to the folder that holds `config.R` and the numbered scripts. The
chunk also confirms each file is present before sourcing anything.

```{r script-dir}
R_DIR <- if (dir.exists("R")) "R" else "."

needed <- c("config.R", "01_preprocess.R", "02_outline_features.R",
            "03_color_features.R", "04_vein_features.R", "05_run_all.R")
present <- file.exists(file.path(R_DIR, needed))
stopifnot("Some scripts are missing from R_DIR — fix R_DIR above." = all(present))
data.frame(script = needed, found = present)
```

# Clear the stale cache

Remove any old `output/features.csv` so there is no chance of knitting the
report against leftover data. (A warning here just means it was already gone.)

```{r clear-cache}
old <- file.path("output", "features.csv")
if (file.exists(old)) {
  file.remove(old)
  message("Removed stale ", old)
} else {
  message("No existing features.csv — nothing to clear.")
}
```

# Source config and confirm the data path

`config.R` defines `DATA_DIR`, `SPECIES_IDS`, the file-name templates, etc.
The check below must print `TRUE` for every species before the slow loop —
this is where a wrong `DATA_DIR` (e.g. a Windows `C:/...` path on a Mac) shows
up immediately instead of 6 minutes into the run.

```{r source-config}
source(file.path(R_DIR, "config.R"))

# DATA_DIR sanity: does the dataset directory exist, and is the first image
# of each requested species reachable?
stopifnot("DATA_DIR does not exist — fix it in config.R" = dir.exists(DATA_DIR))

first_images <- vapply(SPECIES_IDS, function(id) {
  folder <- gsub("\\{id\\}", id, FOLDER_TEMPLATE)
  fname  <- gsub("\\{n\\}", sprintf("%03d", 1),
                 gsub("\\{id\\}", id, FILE_TEMPLATE))
  file.exists(file.path(DATA_DIR, folder, fname))
}, logical(1))

data.frame(species_id = SPECIES_IDS,
           species_name = unname(SPECIES_NAMES[as.character(SPECIES_IDS)]),
           first_image_found = first_images)
```

```{r stop-if-missing}
stopifnot(
  "At least one species' first image was not found — check DATA_DIR and the FILE/FOLDER templates in config.R." =
    all(first_images)
)
message("All ", length(SPECIES_IDS), " species reachable. OK to run the loop.")
```

# Source the feature scripts

These only *define* functions (`preprocess_leaf`, `outline_features`,
`color_features`, `vein_features`). Sourcing them should print package startup
messages but should not process any images.

```{r source-features}
source(file.path(R_DIR, "01_preprocess.R"))
source(file.path(R_DIR, "02_outline_features.R"))
source(file.path(R_DIR, "03_color_features.R"))
source(file.path(R_DIR, "04_vein_features.R"))
```

# Run the full extraction loop

This is the slow step (a few minutes for 5 species at 512px). `05_run_all.R`
runs `run_all()` at the bottom, which writes `output/features.csv` and returns
the data frame as `features`.

```{r run-all}
source(file.path(R_DIR, "05_run_all.R"))
```

# Sanity-check the result

The whole point of this notebook: confirm the CSV has many rows across all
species, not 2 rows of one species.

```{r check-csv}
f <- readr::read_csv(file.path("output", "features.csv"), show_col_types = FALSE)

cat("Total rows:", nrow(f), "\n")
cat("Distinct species:", length(unique(f$species_name)), "\n\n")
print(table(f$species_name))
```

```{r assert-multispecies}
stopifnot(
  "features.csv still has too few rows — the loop did not process the full set." =
    nrow(f) >= 0.5 * length(SPECIES_IDS) * N_PER_SPECIES,
  "features.csv has fewer species than requested in SPECIES_IDS." =
    length(unique(f$species_name)) >= length(SPECIES_IDS)
)
message("features.csv looks good — you can now knit final_report.Rmd.")
```

# Optional: visual check of one leaf

Confirm the threshold cleanly separates leaf from background, and that the
leaf is not rendered mirrored/rotated. Useful for the "threshold quality"
fill-in in the report. Change `id`/`n` to spot-check other species.

```{r visual-check, fig.width=9, fig.height=3.5}
id <- SPECIES_IDS[1]; n <- 1
folder <- gsub("\\{id\\}", id, FOLDER_TEMPLATE)
fname  <- gsub("\\{n\\}", sprintf("%03d", n),
               gsub("\\{id\\}", id, FILE_TEMPLATE))
p <- preprocess_leaf(file.path(DATA_DIR, folder, fname))

par(mfrow = c(1, 3), mar = c(1, 1, 2, 1))
image(t(p$gray[nrow(p$gray):1, ]), col = gray.colors(256), axes = FALSE, main = "grayscale")
image(t(p$mask[nrow(p$mask):1, ]), col = c("white", "black"), axes = FALSE, main = "binary mask")
image(t(p$edge[nrow(p$edge):1, ]), col = c("white", "black"), axes = FALSE, main = "outline")
```

---
